Benchmarking tRNA-Seq quantification approaches by realistic tRNA-Seq data simulation identifies two novel approaches with higher accuracy

Version 1

Abstract

Abstract

Quantification of transfer RNA (tRNA) using illumina sequencing based tRNA-Seq is complicated by their degree of redundancy and extensive modifications. As such, no tRNA-Seq method has become well established, while various approaches have been proposed to quantify tRNAs from sequencing reads. Here, we use realistic tRNA-Seq simulations to benchmark tRNA-Seq quantification approaches, including two novel approaches. We demonstrate that these novel approaches are consistently the most accurate, using data simulated to mimic five different tRNA-Seq methods. This simulation-based benchmarking also identifies specific shortfalls for each quantification approach and suggests that up to 13% of the variance observed between cell lines in real tRNA-Seq data could be due to systematic differences in quantification accuracy.

Introduction

Transfer RNA (tRNA) are abundant, highly structured, and short (70-100 nt) non-coding RNAs which are essential for the conversion of coding sequences into polypeptides1. They can be categorised by their aminoacylation identity into 20 groups, each composed of several tRNAs, known as isoacceptors, that are able to translate synonymous codons with the same amino acid. Adding to this complexity, in higher eukaryotes, there are also isodecoders: tRNAs with the same anticodon but different sequences. tRNAs also undergo substantial chemical modification, containing on average 13 modifications by molecule2 with over 100 post-transcriptional modifications identified to date1,3. Notably, a myriad of human diseases have been associated with dysregulation of tRNA expression4–6 and mutations in tRNA modification enzymes7–10, underscoring their central role in cellular homeostasis.

Given their complexity, quantification of tRNAs is more challenging than other RNA species. Experimentally, tRNAs have been intractable to high-throughput illumina sequencing due to the compact secondary and tertiary structure. In particular, the highly base-paired 5’ and 3’ termini interfere with adaptor ligation11, although this has been addressed in recent publications using splint adapters which take advantage of the 3’ CCA on mature tRNAs11,12. Moreover, some of the post-transcriptional modifications present on tRNAs disrupt the Watson-Crick face and impede reverse transcription (RT), leading to either premature termination of the RT reaction and/or high misincorporation rates around the modified nucleotide13. To alleviate the impact of such modification, demethylating enzymes, such as AlkB, have been utilised in several methods including DM-tRNA-Seq14 and ARM-seq15, although the activity of these enzymes have not been well characterised. An alternative solution is the use of a modification tolerant RT, such as TGIRT14,16, Superscript IV12,17 or marathonRT18, which have a higher processivity though modified nucleotides. Regardless of the approach taken, tRNA-Seq reads inevitably possess increased misincorporation rates and usually 3’ truncations. The processing of tRNA-Seq data is also more complex than other forms of RNA-seq, as the evolution processes exerted upon tRNA genes has led to the accumulation of dozens near identical sequences19. In the case of the human genome and model mammalian species such as mouse and rat, this has generated hundreds of actively transcribed tRNA genes20,21. This sequence redundancy complicates the easy assignment of tRNA-Seq reads to their genomic loci and, therefore, quantification is normally performed at the level of the mature tRNA sequences, or else anticodons.

Multiple quantification approaches for full length illumina-based tRNA-Seq have been employed, all relying upon alignment of reads to tRNA sequences, with considerable variability in how multi-mapped reads are handled. Bowtie222 is the most frequently used aligner for tRNA-Seq12,15,23,24, although aligners designed for small RNA-seq such as SHRiMP25 have also been used11. For tRNA-Seq read alignment with bowtie2, a wide variety of parameterisations have been employed, which frequently include utilising very short and/or non-exact alignment seeds12,23. Due to the high similarity between tRNA sequences, multi-mapped reads present a significant problem, which has again been solved by a variety of means, including the complete exclusion of multi-mapped reads12,14, fractional assignment to all aligned tRNAs17, and random selection from amongst the alignments15. Recently, mimseq, an end-to-end tRNA-Seq quantification pipeline was developed16. This pipeline uses GSNAP26 to align reads to consensus sequences representing clusters of similar tRNAs along with a novel iterative mapping procedure, in which mutated positions are masked. Despite this significant variability between the quantification approaches which have been applied to tRNA-Seq, to date, no systematic comparison of these approaches has been performed.

Here, we used real tRNA-Seq data to generate realistic benchmark data simulations and compare common quantification approaches. This allowed us to identify the best generic quantification approach across different tRNA-Seq methods and to establish specific shortfalls of each approach.

Results

Optimising read alignment for tRNA-Seq

Read alignment software needs to be carefully parameterised for tRNA-Seq and a number of alternative parameterisations have been proposed, predominantly using bowtie2 with a short seed length12,14,17. To evaluate alignment parameters for bowtie2 with a short seed length, we simulated the highest possible quality of tRNA-Seq reads: full length and including only sequencing errors. These reads were then aligned to the reference tRNA sequences with bowtie2, using a sensitive parameterisation, allowing a single error in the seed, as frequently employed for tRNA-Seq to accommodate the increased misincorporation rate12,23. Two parameters were varied: L and D, representing the seed length and how much effort to expend trying to find a better alignment, respectively. This simulation and alignment procedure was performed for tRNA from Homo sapiens and Mus musculus tRNAs separately. We observed that when an error in the seed is allowed, as the seed length is decreased, there needs to be a concomitant increase in effort expended to allow bowtie2 more opportunities to find the best possible alignment, especially with respect to the Transcript ID (Figure 1a). Thus, previously employed alignment parameters have been sub-optimal, even for unrealistically high quality tRNA-Seq reads12.

Alignment parameterisation in the context of frequent multiple-alignment. a). Impact of bowtie2 parameterisation on the percentage of correct alignments. b) The percentage of single and multi-mapped reads at each level of tRNA nomenclature. c). The frequency of reads multi-mapping at the Gene locus ID, Transcript ID and Anticodon level. Frequencies were normalised by dividing by the total number of reads aligned to the two tRNAs. Data shown is from YAMAT-Seq, BT20, replicate A.

Inevitably, tRNA-Seq reads will frequently be mapped to multiple tRNA sequences with equal alignment scores, since some tRNA sequences are nearly identical and real tRNA-Seq reads will be truncated and include misincorporations. To explore the extent of the multi-mapping issue for tRNA-Seq, we aligned reads from ALL-tRNAseq18, DM-tRNA-seq14, mim-tRNAseq16, QuantM-tRNA-seq12 and YAMAT-seq11 samples to the reference tRNA sequences, using bowtie2 (–L 10 –D 100). We then summarised the extent of multi-mapping at four levels using the naming convention from GtRNAdb20: Gene locus ID, Transcript ID, Anticodon and Amino acid. As expected, at the higher levels of tRNA annotation, the number of reads mapping to a single feature increases (Figure 1b). Many pairs of tRNA sequences showed high rates of multi-mapping at the Gene locus ID level, including pairs of tRNAs for different anticodons, but multi-mapping is less prominent at higher levels (Figure 1c). These data demonstrate that multi-mapping is prevalent at the level of the alignments to the tRNA sequences, but it should be less problematic at higher levels if it is dealt with appropriately.

Simulating realistic tRNA-Seq data

To address the question of how to most accurately quantify tRNAs from tRNA-Seq, we aimed to simulate realistic tRNA-Seq data by using the frequency of misincorporations and truncations (hereafter referred to jointly as the ‘error profile’) as observed in tRNA-Seq data. To ensure our findings were generically applicable, error profiles were learned from five tRNA-Seq methods: ALL-tRNAseq, DM-tRNA-seq, mim-tRNAseq, QuantM-tRNA-seq and YAMAT-seq, and two species: Homo sapiens and Mus musculus. This was achieved by first aligning reads to a tRNA reference transcriptome, following which, error profiles were determined from the alignments by tallying mutation frequencies and read alignment coordinates.

Since the mutations and truncations are dependent on the tRNA modifications, the position of the mutations and truncation events in the error profiles should match the positions of known modifications. Furthermore, the relationship between the tRNA modifications and error profiles should depend on whether tRNA modifications are enzymatically removed, which RT used, and whether there was a size selection step prior to sequencing. Thus, samples from different tRNA-Seq methods should have different error profiles. To confirm this, we used the experimentally determined modifications listed in MODOMICS27. As expected, adenosines modified to inosine cause misincorporation of guanine28 (Figure 2a). In contrast, N1-methyladenosine (m1A) frequently causes misincorporation for all methods, except DM-tRNA-seq and ALL-tRNAseq, which employ AlkB to remove methylations. Interestingly, an overall higher proportion of mutations was observed with mim-tRNA-Seq, likely reflecting the optimisation of the RT step to allow more processivity over modified bases (Supplementary Figure 1a). In keeping with the more restrictive size selection of YAMAT-Seq, and the improved RT processivity with mim-tRNA-Seq and ALL-tRNA-Seq, the reads from these methods are less frequently truncated than the other methods (Figure 2b; Supplementary Figure 1b). Notably, while read truncations were frequently close to known modified nucleotides for all methods except YAMAT-Seq, the modified nucleotides around the site of truncation varied considerably between the tRNA-Seq methods (Figure 2c). This is likely due to the considerable differences between the tRNA-Seq library preparation steps for these five methods. Overall, the error profiles appear to reflect the different mutation and truncation rates that would be expected for the five tRNA-Seq methods and should thus enable relatively realistic simulation of tRNA-Seq reads from a broad range of tRNA-Seq protocols.

The observed error profiles reflect the expected misincorporations from modifications in MODOMICS. a). Misincorporation frequencies for Inosine and m1A sites b). The proportion of reads which are truncated and how often this occurs close to a known modification site. c). The proportion of truncated reads which are truncated within 1 nucleotide of common modifications. m2,2G = N2,N2-dimethylguanosine, m2G = N2-methylguanosine, m1Y = 1-methylpseudouridine, m5C = 5-methylcytidine, m1G = 1-methylguanosine, m1A = 1-methyladenosine, D = dihydrouridine, Y = pseudouridine.

From these error profiles, we then simulated 2 datasets: Uniform: an equal number of reads from all tRNA species, with misincorporation and truncations added at the frequencies observed in the real tRNA-Seq data. One set of simulated reads was generated for each tRNA-Seq sample. Realistic: a randomly varied number of reads from each tRNA sequence, drawn from a log2-Gaussian distribution with a mean of 10 and standard deviation of 5, left-censored at zero. Misincorporation and truncations were added at the frequencies observed in the real tRNA-Seq data. Ten sets of simulated reads were generated for each tRNA-Seq sample.

Comparing alignment strategies

The first question we wished to address was which read alignment strategy yielded the most accurate alignments. We therefore compared 3 strategies using the Uniform simulation data by employing either bowtie2 or SHRiMP alignment to tRNA sequences or alignment to consensus tRNA sequences using GSNAP, using the mimseq pipeline16.

Bowtie2 and SHRiMP alignments were analysed at 3 levels using the naming convention from GtRNAdb: Gene locus ID, Transcript ID and Anticodon. The mimseq pipeline aligns to consensus tRNA sequences that approximate a single transcript ID and are associated with a single anticodon. Therefore, we also analysed the bowtie2 and SHRiMP alignments at the ‘Mimseq isodecoder’ level to make them comparable.

With the configurations used here, bowtie2 and SHRiMP aligned the most reads, with the median percentage of aligned reads across all Homo sapiens samples being 98.5%, 93.0% and 71.3%, for bowtie2, SHRiMP and mimseq (GSNAP), respectively (Figure 3a). However, relative to mimseq, SHRiMP and bowtie2 had up to 3.3% and 4.0%, respectively, fewer alignments correct at the anticodon level for Homo sapiens samples (Figure 3b; supplementary Figure 2a). This suggests there is an expected trade-off between the proportion of aligned reads and the accuracy of the alignments (Supplementary Figure 2b).

Alignment accuracy is dependent upon the alignment strategy and tRNA-Seq method. a) Percentage of reads aligned. b). Percentage of correct alignments. c). Spearman correlations for read misalignments rates between pairs of tRNA-Seq methods or pairs of samples within a single tRNA-Seq method. QuantM-tRNA-seq is not included since these samples were only for Mus musculus. d). A comparison of correct assignments for 4 selected Homo sapiens anticodons which are most variable across tRNA-Seq methods. QuantM-tRNA-seq is not included since these samples were only for Mus musculus.

We next considered the consistency for the read misalignments patterns. Notably, the read misalignments were more correlated between samples from the same tRNA-Seq method than between the two tRNA-Seq methods for each species (Figure 3c), likely due to the differences in error profiles between each tRNA-Seq method (Figure 1). Interestingly, some tRNAs anticodons were difficult to align when simulating reads from specific tRNA-Seq methods. For example, using bowtie2, reads from Homo sapiens Sup-TTA tRNAs are more correctly aligned for YAMAT-seq than DM-tRNA-seq. In contrast, mimseq accurately aligned Sup-TTA reads simulated from DM-tRNA-seq, but not from YAMAT-seq (Figure 3d).

We also observed cell-line specific misalignment patterns that vary across the tRNA-Seq methods and alignment approaches (Supplementary Figure 2c). For example, using mimseq, reads simulated from mim-tRNAseq for Asp-ATC tRNAs were missassigned to Cys-GCA tRNAs 67%, 40% and 100% of the time for BT20, MCF7, and SKBR3, respectively. However, using bowtie2, the misalignment rates were only 18%, 24% and 36%, respectively. The difference in the misalignments between the cell lines and tRNA-Seq methods supports the approach taken herein to use learn sample-specific error profiles from multiple tRNA-Seq methods to enable simulation of a broad set of representative samples from which to identify a generically accurate quantification approach.

Handling multiple mapped reads

Once reads are aligned, the major variable is how to handle multi-mapped reads in the process of tallying read counts per tRNA. To compare approaches, we aligned the Realistic simulated reads to tRNA sequences with bowtie2 or SHRiMP and applied 4 approaches used previously to count reads per tRNA, alongside two further approaches which we expected would be more accurate that have not previously been applied, to our knowledge (Table 1). In addition, we used the dedicated tRNA-Seq quantification tool, mimseq.

With the exception of Decision and mimseq, which inherently generate quantification at multiple levels, all other approaches quantify at the level of the Gene locus ID sequences included in the reference fasta. We further summarised quantification to transcript ID and anticodon-level feature abundances by summation. We also summarised the quantification values to the same ‘mimseq isodecoder’ features quantified by mimseq, to make them comparable. A summary of the 4 levels of quantification and how these are obtained for each read tallying approach is shown (Table 2).

The proportion of reads used by each read tallying approach varies considerably. Salmon, Fractional and Random use all mapped reads for each level of quantification, whereas stricter approaches which filter by multimapping or mapping quality use the least reads and frequently assign fewer than 50% of the mapped reads (Figure 4a; Supplementary Figure 3a). As intended, Decision uses increasingly more reads at higher levels of quantification. Importantly, Decision uses approximately the same number of reads as mimseq at the mimseq isodecoder level, but consistently more at the anticodon level, where it approaches the use of all mapped reads. In contrast, Mimseq uses the same number of reads for both mimseq isodecoder and anticodon level quantification. This suggests Decision is more appropriately determining which reads to utilise for each level of quantification.

a) Reads assigned at anticodon and mimseq isodecoder level for each read tally method. Only results from alignment with bowtie2 or GSNAP (Mimseq) are shown. b) Mean RMSE at anticodon and mimseq isodecoder level. *=Best approach for each tRNA-Seq method, ▾=Worst approach. c). As per (b), except Pearson correlation coefficient. d) The percentage of reads from each simulation sample which are single or multiple aligned, separated by whether the alignment position(s) have the correct anticodon. Multi-aligned reads with more than more anticodon were deemed incorrect.

Comparing read tallying approaches

To assess the overall accuracy of the abundance estimates obtained with the varying read tallying approaches, we calculated two metrics. The root mean squared error (RMSE) is a common error metric which captures how closely the quantification estimates match the ground truth29. However, in many applications of tRNA-Seq, including the comparison between conditions or tissues, it’s sufficient to accurately capture the abundance relationship between samples, even if there is a systematic over or under-estimation across all samples. Hence, we complemented the evaluation using RMSE by also considering the Pearson correlation coefficient between the quantification estimates and ground truth.

To compare the quantification approaches, we considered the mean RMSE and mean correlation across all features at a given quantification level. At the anticodon and mimseq isodecoder level, our novel approaches, Decision and Salmon consistently show the lowest RMSE and highest correlation, suggesting they deal with multi-mapped reads most appropriately. Of the previously used read-tallying approaches, Mimseq shows the highest accuracy. Surprisingly, Random and Fractional also perform well at the anticodon level, while Unique and MAPQ>10 performed poorly (Figure 4b-c). On further inspection, this is because the multi-mapped reads are predominantly mapped to multiple tRNA species from the same anticodon (Figure 4d; Supplementary Figure 3a). Thus, requiring unique alignments at the level of individual tRNA sequences (using Unique or MAPQ>10) leads to detrimental discarding of multi-mapped reads which could be assigned to a single anticodon. For these reads, selecting one alignment at random, or equally distributing the read between all of them, is a more sensible approach and thus leads to more accurate anticodon level quantification. However, for quantification at lower levels, Unique and MAPQ > 10 are more accurate than Random and Fractional, suggesting discarding of multi-mapped reads is better than naive resolution for quantification below the anticodon level (Figure 4b-c). Very similar results are observed when SHRiMP is used in place of bowtie2 (Supplementary Figure 3b-c). Finally, for quantification at the level of individual tRNA sequences, all methods perform equally well, with the exception of Random, which performed the worst. Overall, Decision and Salmon are consistently the best read tallying methods to apply.

We reasoned that the misalignment of reads should be one of the major factors reducing quantification accuracy for all methods. In support of this, the fraction of reads correctly aligned and the quantification evaluation metrics were correlated (Supplementary Figure 4a-b). The exception was MAPQ>10 at anticodon level, which shows poor accuracy even when the proportion of correctly aligned reads is high. This agrees with the previous observation that discarding low quality reads is detrimental for anticodon-level read tallying because reads are typically mapped across multiple tRNAs for the same anticodon, not between anticodons (Figure 3d). Thus, excepting quantification by MAPQ>10, our results indicate that misalignment is one of the major issues with tRNA-Seq quantification.

Using simulations to inform interpretation of results

Summarising the RMSE and correlation metrics to the mean value over all features enables read tallying approaches to be easily compared, but masks fine-grained differences which go against the trend and could give false confidence when interpreting quantification of individual features. We therefore further considered which features each approach quantifies best, focusing on the simulations from the YAMAT-Seq data and quantification at the anticodon level. For these simulations, Decision had the highest mean correlation and lowest mean RMSE, with Salmon performing nearly equivalently and Mimseq also being accurate (Figure 4c). However, we observed that the correlations for some anticodons were very sensitive to the read tallying approach (Figure 5a-b). For example, the correlation for Ser-GGA is much lower with Mimseq than Decision or Salmon, whilst the correlation for Pro-AGG is much higher with Mimseq than Decision or Salmon (Figure 5b; Supplementary Figure 4c). This suggests the choice of quantification approach may have different impacts on quantification accuracy for each anticodon and the interpretation of results should ideally take into account the limits of each quantification approach.

Each quantification approach has specific shortfalls. a) Pearson correlation between ground truth and estimated read counts for six anticodons with variable quantification accuracy between read tallying approaches. b). c) The variance in observed fold-changes in the real tRNA-Seq data that is explained by fold-changes in the uniform simulation data. x = The mean variance explained across the four Homo sapiens datasets.

Given the observed association between read misassignment and quantification accuracy, we hypothesised that differences in read misalignments between cell lines could erroneously induce apparent differences in tRNA feature abundances. To test this, we compared the fold-change between cell lines when simulating the same number of reads for each tRNA with the fold-changes observed in the real data. The fold-changes with the simulated data should reflect technical biases resulting from systematic differences in misassignments between cell lines due to differences in the error profiles. Strikingly, up to 13% of the variance in fold-changes in the real data were explained by technical biases (Figure 5c). Even using the most accurate quantification method, Decision, up to 7% of the variance for anticodon level fold-changes in the real data is explained by technical biases. All quantification methods appeared to show some propensity to identify fold-changes between conditions that were partially explained by technical biases for at least one level of tRNA quantification. The exception to this was salmon, for which a maximum of 2% of the variance in fold-changes were explained by technical biases. This suggests salmon may be a more trustable quantification method for the detection of fold-changes between experimental conditions.

Discussion

Recent advances in sequencing using nanopores promise a potential revolution for RNA sequencing, including tRNAs30. Despite this, illumina sequencing is expected to remain the dominant approach for tRNA sequencing, at least in the short-term. The first tRNA-seq approaches using illumina sequencing were published in 201514,15, since which, the comparison of bioinformatic approaches for tRNA quantification from tRNA-Seq has received surprisingly little attention. In part, this may be because tRNA-Seq methods continue to be developed and the optimal quantification is considered to be dependent upon the method. Encouragingly, however, we have identified two robust quantification methods that appear to be suitable for all tRNA-Seq data.

An important aspect of our benchmarking is the simulation of realistic tRNA-Seq data, which has not been attempted before, to our knowledge. The only previous attempt to simulate tRNA-Seq reads that we are aware of did not capture the effect of modifications on read truncations and mutations31. By learning error profiles of individual samples and then simulating data from these error profiles, our simulations at least partially capture the technical variability between samples, cell lines and tRNA-Seq methods. However, there are limitations to this approach. All truncation and misincorporation events are measured independently, such that the error profiles will not capture dependencies between the events, which occur with respect to correlated misincorporations at nucleotides due to linked modifications32. In addition, the error profiles are inherently biased by the ability to align the tRNA-Seq reads. Misincorporations and truncations which prevent the read being aligned appropriately will not be captured in the error profiles. As such, the simulated data may be considered to simulate the alignable portion of a tRNA-Seq sample, with biases due to the alignment strategy used. Here, we attempted to mitigate this by aligning reads with multiple aligners. Despite these limitations, our simulations represent the most realistic simulation of tRNA-Seq data to date, and enabled us to perform a rigorous comparison of tRNA-Seq quantification approaches.

The crucial steps for accurate tRNA-Seq quantification are the alignment of reads and handling of multi-mapped reads, with the latter being the step which varies most between the approaches that have been applied to date. Although reads may be multi-mapped across different tRNA species, we show here that the issue is most pressing at the Gene locus ID and Transcript ID level, and much less so at the Anticodon level. However, dealing with multi-mapped reads inappropriately can itself create issues at the anticodon level. The approaches used to date resolve the multi-mapping are only suitable for quantification at particular levels, which has not previously been appreciated. Crucially, the common approach of removing multi-mapped reads and/or low quality reads is significantly suboptimal for the anticodon level quantification which it has been previously employed for12,14, since it leads to the unnecessary discarding of reads which have multi-mapped across tRNAs with the same anticodon. The bespoke quantification approach of Mimseq is better than filtering multi-mapped reads, but our simulations indicate it performed no better than simple random assignment of multi-mapped reads for anticodon-level quantification.

The central observation of this benchmarking study is that two simple approaches alleviate the issue of multi-mapped reads and perform better than existing approaches at all levels of quantification. This is observed across all 4 of the tRNA-Seq methods, suggesting our observation can likely be extrapolated to all tRNA-Seq methods using illumina sequencing for full length tRNAs. The Salmon-based approach involves reporting all alignment positions and then using salmon to probabilistically resolve the multi-mapped reads. The Decision approach involves counting reads at each level of quantification where the read maps to tRNAs from a single feature, e.g multiple tRNAs from the same anticodon. Given the lower propensity for technical biases to explain fold-changes in real data with Salmon and the reduced barrier to its use, we suspect salmon will prove to be the most popular. For this approach, bowtie2 simply needs to be run with the –a flag to report all reads. The alignments must then be filtered to retain only the equal top-scoring alignments for each read, before passing the alignments to salmon quant.

While the motivation of this study was to determine the optimal overall quantification approach, our simulations also identified specific tRNAs that cannot be accurately quantified in particular samples. Our simulations further allowed us to establish that a small proportion of the observed variability between cell lines may not be biological. Thus, comparisons between samples should be interpreted with caution, since they may be partially due to technical biases in read miss-alignment. We propose that a simulation-based approach could be applied more widely as a quality control step to highlight features whose quantification may be inaccurate. In particular, where changes in tRNA abundance between cell lines are observed, it would be prudent to use similar simulations to demonstrate that such changes are not driven by differences in tRNA modifications driving systematic differences in misincorporations and truncations.

The decision about what level to interrogate tRNA abundance is a trade off between a desire to obtain the most ‘fine-grained’ data possible, and the practical consideration of how accurately tRNAs can be quantified at each level. tRNA isodecoders have been found to be functionality distinct33,34 and differential abundance of specific Transcript IDs has been identified across mouse tissues12 and human cell lines16. Nonetheless, tRNA abundances are more accurately quantified at the Anticodon level and this is likely to remain the predominant approach for the foreseeable future. As tRNA-Seq methods continue to improve the processivity of reverse transcriptase over modified bases and yield fewer truncated reads, the potential for highly accurate Transcript ID level quantification will open up. Sequencing tRNAs using nanopores is also likely to provide a significant benefit in this respect30. However, these improvements in experimental methods need to be complemented by rigorous improvements in the data processing to take full advantage of the potential benefits. We envisage the simulation-based approach we have used here being reutilised as new experimental methods are developed, to maintain a high standard of tRNA-Seq data analysis.

Data and code availability

The data processing pipeline is available at https://github.com/MRCToxBioinformatics/pipeline_compare_trnaseq_quant and the version used (v0.1) is archived at zenodo, DOI: 10.5281/zenodo.10229154. The pipeline makes extensive use of a dedicated python package we developed for the simulation of tRNA-Seq samples, simulatetrna, which is available at https://github.com/MRCToxBioinformatics/simulate_trna and the version used (v0.1) is archived at zenodo: DOI: 10.5281/zenodo.10235438. The downstream data analysis and visualisation was performed using R markdown notebooks 35, which are available at https://github.com/MRCToxBioinformatics/trna_seq_quantification_benchmarking (v0.1.1), archived with zenodo, DOI: 10.5281/zenodo.10372994.

eLife Assessment

This study provides a valuable resource by thoroughly benchmarking multiple sequencing-based tRNA quantification methods. The suggested best practice is supported by solid evidence from in silico experiments in multiple scenarios. The major weakness of the manuscript is the incomplete validation of newly generated experimental datasets.

Reviewer #1 (Public Review):

Summary:

In the manuscript titled "Benchmarking tRNA-Seq quantification approaches by realistic tRNA-Seq data simulation identifies two novel approaches with higher accuracy," Tom Smith and colleagues conducted a comparative evaluation of various sequencing-based tRNA quantification methods. The inherent challenges in accurately quantifying tRNA transcriptional levels, stemming from their short sequences (70-100nt), extensive redundancy (~600 copies in human genomes with numerous isoacceptors and isodecoders), and potential for over 100 post-transcriptional chemical modifications, necessitate sophisticated approaches. Several wet-experimental methods (QuantM-tRNA, mim-tRNA, YAMAT, DM-tRNA, and ALL-tRNA) combined with bioinformatics tools (bowtie2-based, SHRiMP, and mimseq) have been proposed for this purpose. However, their practical strengths and weaknesses have not been comprehensively explored to date. In this study, the authors systematically assessed and compared these methods, considering factors such as incorrect alignments, multiple alignments, misincorporated bases (experimental errors), truncated reads, and correct assignments. Additionally, the authors introduced their own bioinformatic approaches (referred to as Decision and Salmon), which, while not without flaws (as perfection is unattainable), exhibit significant improvements over existing methods.

Strengths:

The manuscript meticulously compares tRNA quantification methods, offering a comprehensive exploration of each method's relative performance using standardized evaluation criteria. Recognizing the absence of "ground-truth" data, the authors generated in silico datasets mirroring common error profiles observed in real tRNA-seq data. Through the utilization of these datasets, the authors gained insights into prevalent sources of tRNA read misalignment and their implications for accurate quantification. Lastly, the authors proposed their downstream analysis pipelines (Salmon and Decision), enhancing the manuscript's utility.

Weaknesses:

As discussed in the manuscript, the error profiles derived from real-world tRNA-seq datasets may still harbor biases, as reads that failed to "align" in the analysis pipelines were not considered. Additionally, the authors did not validate the efficacy of their "best practice" pipelines on new real-world datasets, preferably those generated by the authors themselves. Such validation would not only confirm the improvements but also demonstrate how these pipelines could alter biological interpretations.

Because tRNA-sequencing methods have not been widely used (compared to mRNA-seq), many readers would not be familiar with the characteristics of different methods introduced in this study (QuantM-tRNA, mim-tRNA, YAMAT, DM-tRNA, and ALL-tRNA; bowtie2-based, SHRiMP, and mimseq; what are the main features of "Salmon?"). The manuscript will read better when the basic features of these methods are described in the manuscript, however brief.

Reviewer #2 (Public Review):

Summary:

The authors provided benchmarking study results on tRNA-seq in terms of read alignment and quantification software with optimal parameterization. This result can be a useful guideline for choosing optimal parameters for tRNA-seq read alignment and quantification.

Strengths:

Benchmarking results for read alignment can be a useful guideline to choose optimal parameters and mapping strategy (mapping to amino acid) for various tRNAseq.

Weaknesses:

The topic is highly specific, and the novelty of the analysis might not be widely useful for general readers.

Some details of the sequencing data analysis pipeline are not clear for general readers:

(1) The explanation of the parameter D for bowtie2 sounds ambiguous. "How much effort to expend" needs to be explained in more detail.

(2) Please provide optimal parameters (L and D) for tRNA-seq alignment.

(3) I think the authors chose L=10 and D=100 based on Figure 1A. Which dataset did you choose for this parameterization among ALL-tRNAseq, DM-tRNAseq, mim-tRNAseq, QuantM-tRNA-seq, and YAMAT-seq?

(4) Salmon does not need a read alignment process such as Bowtie2. Hence, it is not clear "Only results from alignment with bowtie2" in Figure legend for Figure 4a.

Author response:

The following is the authors’ response to the original reviews.

Reviewer 1:

Because tRNA-sequencing methods have not been widely used (compared to mRNA-seq), many readers would not be familiar with the characteristics of different methods introduced in this study (QuantM-tRNA, mim-tRNA, YAMAT, DM-tRNA, and ALL-tRNA; bowtie2-based, SHRiMP, and mimseq; what are the main features of "Salmon?"). The manuscript will read better when the basic features of these methods are described in the manuscript, however brief.

Introduction page 4 now clarifies a little more the difference between bowtie2, SHRiMP and mimseq. Results page 9 briefly summarises the differences between the tRNA-Seq methods. Results page 14 clarifies how Decision and Salmon work.

Reviewer 2:

(1) The explanation of the parameter D for bowtie2 sounds ambiguous. "How much effort to expend" needs to be explained in more detail.

Results page 6 gives a more precise explanation of the D parameter.

(2) Please provide optimal parameters (L and D) for tRNA-seq alignment.

I think optimal here is not possible to determine. It will depend on the species, the frequency of misincorporations due to modifications (tRNA-Seq protocol specific) and how long one is willing to let bowtie continue searching for a better match. The point of Figure 1a is that D needs to be increased if L is decreased and an error is allowed in the seed. I think the sentence in the results section Figure 1a is the appropriate way to express this without committing to a single ‘optimal’ parameterisation_:_ ‘We observed that when an error in the seed is allowed, as the seed length is decreased, there needs to be a concomitant increase in effort expended to allow bowtie2 more opportunities to find the best possible alignment, especially with respect to the Transcript ID‘.

(3) I think the authors chose L=10 and D=100 based on Figure 1A. Which dataset did you choose for this parameterization among ALL-tRNAseq, DM-tRNAseq, mim-tRNAseq, QuantM-tRNA-seq, and YAMAT-seq?

Figure 1A is based on simulation of full length reads with only sequencing errors, e.g not from any tRNA-Seq method in particular. This is stated in the results text and I’ve clarified in the figure legend.

(4) Salmon does not need a read alignment process such as Bowtie2. Hence, it is not clear "Only results from alignment with bowtie2" in Figure legend for Figure 4a.

I’m using Salmon in ‘alignment-mode’, taking the alignments from bowtie2. I’ve clarified this in results page 14.

Version 2

Abstract

Abstract

Quantification of transfer RNA (tRNA) using illumina sequencing based tRNA-Seq is complicated due to their degree of redundancy and extensive modifications. As such, no tRNA-Seq method has become well established, while various approaches have been proposed to quantify tRNAs from sequencing reads. Here, we use realistic tRNA-Seq simulations to benchmark tRNA-Seq quantification approaches, including two novel approaches. We demonstrate that these novel approaches are consistently the most accurate, using data simulated to mimic five different tRNA-Seq methods. This simulation-based benchmarking also identifies specific shortfalls for each quantification approach and suggests that up to 13% of the variance observed between cell lines in real tRNA-Seq data could be due to systematic differences in quantification accuracy.

Introduction

Transfer RNA (tRNA) are abundant, highly structured, short (70-100 nt) non-coding RNAs that are essential for the conversion of coding sequences into polypeptides1. They can be categorised by their aminoacylation identity into 20 groups, each composed of several tRNAs, known as isoacceptors, that are able to translate synonymous codons with the same amino acid. Adding to this complexity, in higher eukaryotes, there are also isodecoders: tRNAs with the same anticodon but different sequences. tRNAs also undergo substantial chemical modification, containing on average 13 modifications by molecule2 with over 100 post-transcriptional modifications identified to date1,3. Notably, a myriad of human diseases have been associated with dysregulation of tRNA expression4–6 and mutations in tRNA modification enzymes7–10, underscoring their central role in cellular homeostasis.

Given their complexity, quantification of tRNAs is more challenging than other RNA species. Experimentally, tRNAs have been intractable to high-throughput illumina sequencing due to the compact secondary and tertiary structure. In particular, the highly base-paired 5’ and 3’ termini interfere with adaptor ligation11, although this has been addressed in recent publications using splint adapters which take advantage of the 3’ CCA on mature tRNAs11,12. Moreover, some of the post-transcriptional modifications present on tRNAs disrupt the Watson-Crick face and impede reverse transcription (RT), leading to either premature termination of the RT reaction and/or high misincorporation rates around the modified nucleotide13. To alleviate the impact of such modification, demethylating enzymes, such as AlkB, have been utilised in several methods including DM-tRNA-Seq14 and ARM-seq15, although the activity of these enzymes have not been well characterised. An alternative solution is the use of a modification tolerant RT, such as TGIRT14,16, Superscript IV12,17 or marathonRT18, which have a higher processivity though modified nucleotides. Regardless of the approach taken, tRNA-Seq reads inevitably possess increased misincorporation rates and usually 3’ truncations. The processing of tRNA-Seq data is also more complex than other forms of RNA-seq, as the evolution processes exerted upon tRNA genes has led to the accumulation of dozens near identical sequences19. In the case of the human genome and model mammalian species such as mouse and rat, this has generated hundreds of actively transcribed tRNA genes20,21. This sequence redundancy complicates the easy assignment of tRNA-Seq reads to their genomic loci and, therefore, quantification is normally performed at the level of the mature tRNA sequences, or else anticodons.

Multiple quantification approaches for full length illumina-based tRNA-Seq have been employed, all relying upon alignment of reads to tRNA sequences, with considerable variability in how multi-mapped reads are handled. Bowtie2 was designed to align reads accurately and quickly22 and is the most frequently used aligner for tRNA-Seq12,15,23,24. A variety of bowtie2 parameterisations employed to alleviate the issue of misincorporation at modified bases, primarily the use of very short and/or non-exact alignment seeds, which ensures the second stage of alignment by dynamic programming has at least a minimal alignment seed from which to extend12,23. Aligners designed for small RNA-seq such as SHRiMP25 have also been used11. SHRiMP performs a more exhaustive search for the correct alignment and handles high error rates well, which likely makes it suitable for tRNA-Seq alignment with default parameterisation, albeit at a cost of increased runtime and memory requirements25,26.

Due to the high similarity between tRNA sequences, multi-mapped reads present another significant problem, which has again been solved by a variety of means, including the complete exclusion of multi-mapped reads12,14, fractional assignment to all aligned tRNAs17, and random selection from amongst the alignments15. Recently, mimseq, an end-to-end tRNA-Seq quantification pipeline was developed, which proposes a novel approach to deal with modified bases and multi-mapped reads16. This pipeline uses GSNAP27 to align reads to consensus sequences representing clusters of similar tRNAs along with an iterative mapping procedure, in which mutated positions are masked. Despite this significant variability between the quantification approaches which have been applied to tRNA-Seq, to date, no systematic comparison of these approaches has been performed.

Here, we used real tRNA-Seq data to generate realistic benchmark data simulations and compare common quantification approaches. This allowed us to identify the best generic quantification approach across different tRNA-Seq methods and to establish specific shortfalls of each approach.

Results

Optimising read alignment for tRNA-Seq

Read alignment software needs to be carefully parameterised for tRNA-Seq and a number of alternative parameterisations have been proposed, predominantly using bowtie2 with a short seed length12,14,17. To evaluate alignment parameters for bowtie2 with a short seed length, we simulated the highest possible quality of tRNA-Seq reads: full length and including only sequencing errors. These reads were then aligned to the reference tRNA sequences with bowtie2, using a sensitive parameterisation, allowing a single error in the seed, as frequently employed for tRNA-Seq to accommodate the increased misincorporation rate12,23. Two parameters were varied: L and D, representing the seed length and how many seed extensions in a row can fail to produce an improved alignment before the search is halted, respectively.

This simulation and alignment procedure was performed for tRNA from Homo sapiens and Mus musculus tRNAs separately. We observed that when an error in the seed is allowed, as the seed length is decreased, there needs to be a concomitant increase in effort expended to allow bowtie2 more opportunities to find the best possible alignment, especially with respect to the Transcript ID (Figure 1a). Thus, previously employed alignment parameters have been sub-optimal, even for unrealistically high quality tRNA-Seq reads12.

Alignment parameterisation in the context of frequent multiple-alignment

a) Impact of bowtie2 parameterisation on the percentage of correct alignments. Reads were simulated as full length and with only sequencing errors. b) The percentage of single and multi-mapped reads at each level of tRNA nomenclature. c) The frequency of reads multi-mapping at the Gene locus ID, Transcript ID and Anticodon level. Frequencies were normalised by dividing by the total number of reads aligned to the two tRNAs. Data shown is from YAMAT-Seq, BT20, replicate A.

Inevitably, tRNA-Seq reads will frequently be mapped to multiple tRNA sequences with equal alignment scores, since some tRNA sequences are nearly identical and real tRNA-Seq reads will be truncated and include misincorporations. To explore the extent of the multi-mapping issue for tRNA-Seq, we aligned reads from ALL-tRNAseq18, DM-tRNA-seq14, mim-tRNAseq16,

QuantM-tRNA-seq12 and YAMAT-seq11 samples to the reference tRNA sequences, using bowtie2 (-L 10 -D 100). We then summarised the extent of multi-mapping at four levels using the naming convention from GtRNAdb20: Gene locus ID, Transcript ID, Anticodon and Amino acid.

As expected, at the higher levels of tRNA annotation, the number of reads mapping to a single feature increases (Figure 1b). Many pairs of tRNA sequences showed high rates of multi-mapping at the Gene locus ID level, including pairs of tRNAs for different anticodons, but multi-mapping is less prominent at higher levels (Figure 1c). These data demonstrate that multi-mapping is prevalent at the level of the alignments to the tRNA sequences, but it should be less problematic at higher levels if it is dealt with appropriately.

Simulating realistic tRNA-Seq data

To address the question of how to most accurately quantify tRNAs from tRNA-Seq, we aimed to simulate realistic tRNA-Seq data by using the frequency of misincorporations and truncations (hereafter referred to jointly as the ‘error profile’) as observed in tRNA-Seq data. To ensure our findings were generically applicable, error profiles were learned from five tRNA-Seq methods: ALL-tRNAseq, DM-tRNA-seq, mim-tRNAseq, QuantM-tRNA-seq and YAMAT-seq, and two species: Homo sapiens and Mus musculus. These methods include various sequencing library preparation modifications specifically designed to handle tRNAs, primarily the use of the demethylating enzyme AlkB (ALL-tRNA-seq and DM-tRNA-seq) and/or modification tolerant RTs (ALL-tRNAseq, DM-tRNA-seq, mim-tRNAseq and QuantM-tRNA-seq), as well as the use of mature tRNA-specific double-stranded adapters (YAMAT-seq and QuantM-tRNA-seq). The error profiles obtained for each tRNA-Seq method should reflect the specific library preparation steps employed and a quantification approach which works across data simulated from all these methods should be generally applicable to tRNA-Seq. Reads were aligned to a tRNA reference transcriptome, following which, error profiles were determined from the alignments by tallying mutation frequencies and read alignment coordinates.

Since the mutations and truncations are dependent on the tRNA modifications, the position of the mutations and truncation events in the error profiles should match the positions of known modifications. Furthermore, the relationship between the tRNA modifications and error profiles should depend on whether tRNA modifications are enzymatically removed, which RT is used, and whether there was a size selection step prior to sequencing. To confirm this, we used the experimentally determined modifications listed in MODOMICS28. As expected, adenosines modified to inosine cause misincorporation of guanine29 (Figure 2a). In contrast, N1-methyladenosine (m1A) frequently causes misincorporation for all methods, except DM-tRNA-seq and ALL-tRNAseq, which employ AlkB to remove methylations. Interestingly, an overall higher proportion of mutations was observed with mim-tRNA-Seq, likely reflecting the optimisation of the RT step to allow more processivity over modified bases (Supplementary Figure 1a). In keeping with the more restrictive size selection of YAMAT-Seq, and the improved RT processivity with mim-tRNA-Seq and ALL-tRNA-Seq, the reads from these methods are less frequently truncated than the other methods (Figure 2b; Supplementary Figure 1b). Notably, while read truncations were frequently close to known modified nucleotides for all methods except YAMAT-Seq, the modified nucleotides around the site of truncation varied considerably between the tRNA-Seq methods (Figure 2c). This is likely due to the considerable differences between the tRNA-Seq library preparation steps for these five methods. Overall, the error profiles appear to reflect the different mutation and truncation rates that would be expected for the five tRNA-Seq methods and should thus enable relatively realistic simulation of tRNA-Seq reads from a broad range of tRNA-Seq protocols.

The observed error profiles reflect the expected misincorporations from modifications in MODOMICS

a). Misincorporation frequencies for Inosine and m1A sites b). The proportion of reads which are truncated and how often this occurs close to a known modification site. c). The proportion of truncated reads which are truncated within 1 nucleotide of common modifications. m2,2G = N2,N2-dimethylguanosine, m2G = N2-methylguanosine, m1Y = 1-methylpseudouridine, m5C = 5-methylcytidine, m1G = 1-methylguanosine, m1A = 1-methyladenosine, D = dihydrouridine, Y = pseudouridine.

From these error profiles, we then simulated 2 datasets: Uniform: an equal number of reads from all tRNA species, with misincorporation and truncations added at the frequencies observed in the real tRNA-Seq data. One set of simulated reads was generated for each tRNA-Seq sample. Realistic: a randomly varied number of reads from each tRNA sequence, drawn from a log2-Gaussian distribution with a mean of 10 and standard deviation of 5, left-censored at zero. Misincorporation and truncations were added at the frequencies observed in the real tRNA-Seq data. Ten sets of simulated reads were generated for each tRNA-Seq sample.

Comparing alignment strategies

The first question we wished to address was which read alignment strategy yielded the most accurate alignments. We therefore compared 3 strategies using the Uniform simulation data by employing either bowtie2 or SHRiMP alignment to tRNA sequences or alignment to consensus tRNA sequences using GSNAP, using the mimseq pipeline16. Bowtie2 and SHRiMP alignments were analysed at 3 levels using the naming convention from GtRNAdb: Gene locus ID, Transcript ID and Anticodon. The mimseq pipeline aligns to consensus tRNA sequences that approximate a single transcript ID and are associated with a single anticodon. Therefore, we also analysed the bowtie2 and SHRiMP alignments at the ‘Mimseq isodecoder’ level to make them comparable.

With the configurations used here, bowtie2 and SHRiMP aligned the most reads, with the median percentage of aligned reads across all Homo sapiens samples being 98.5%, 93.0% and 71.3%, for bowtie2, SHRiMP and mimseq (GSNAP), respectively (Figure 3a). However, relative to mimseq, SHRiMP and bowtie2 had up to 3.3% and 4.0%, respectively, fewer alignments correct at the anticodon level for Homo sapiens samples (Figure 3b; supplementary Figure 2a). This suggests there is an expected trade-off between the proportion of aligned reads and the accuracy of the alignments (Supplementary Figure 2b).

Alignment accuracy is dependent upon the alignment strategy and tRNA-Seq method

a) Percentage of reads aligned. b). Percentage of correct alignments. c). Spearman correlations for read misalignments rates between pairs of tRNA-Seq methods or pairs of samples within a single tRNA-Seq method. QuantM-tRNA-seq is not included since these samples were only for Mus musculus. d). A comparison of correct assignments for 4 selected Homo sapiens anticodons which are most variable across tRNA-Seq methods. QuantM-tRNA-seq is not included since these samples were only for Mus musculus.

We next considered the consistency for the read misalignments patterns. Notably, the read misalignments were more correlated between samples from the same tRNA-Seq method than between the two tRNA-Seq methods for each species (Figure 3c), likely due to the differences in error profiles between each tRNA-Seq method (Figure 1). Interestingly, some tRNAs anticodons were difficult to align when simulating reads from specific tRNA-Seq methods. For example, using bowtie2, reads from Homo sapiens Sup-TTA tRNAs are more correctly aligned for YAMAT-seq than DM-tRNA-seq. In contrast, mimseq accurately aligned Sup-TTA reads simulated from DM-tRNA-seq, but not from YAMAT-seq (Figure 3d).

We also observed cell-line specific misalignment patterns that vary across the tRNA-Seq methods and alignment approaches (Supplementary Figure 2c). For example, using mimseq, reads simulated from mim-tRNAseq for Asp-ATC tRNAs were missassigned to Cys-GCA tRNAs 67%, 40% and 100% of the time for BT20, MCF7, and SKBR3, respectively. However, using bowtie2, the misalignment rates were only 18%, 24% and 36%, respectively. The difference in the misalignments between the cell lines and tRNA-Seq methods supports the approach taken herein to use learn sample-specific error profiles from multiple tRNA-Seq methods to enable simulation of a broad set of representative samples from which to identify a generically accurate quantification approach.

Handling multiple mapped reads

Once reads are aligned, the major variable is how to handle multi-mapped reads in the process of tallying read counts per tRNA. To compare approaches, we aligned the Realistic simulated reads to tRNA sequences with bowtie2 or SHRiMP and applied 4 approaches used previously to count reads per tRNA, alongside two further approaches which we expected would be more accurate that have not previously been applied, to our knowledge (Table 1).The Decision approach only assigns reads to a tRNA if they are uniquely mapped, but re-frames the uniqueness of the mapping with respect to the level of tRNA quantification being performed. For example, a read aligning to multiple tRNA sequences with the same transcript ID would be assigned to that transcript ID and the associated Anticodon, but would be discarded for Gene locus ID quantification. Salmon implements a lightweight mapping procedure and a maximum-likelihood approach to estimate RNA abundances from the mapped reads, while accommodating multi-mapped reads30. To ensure the quantification approaches were comparable, we used Salmon in the ‘alignment-based’ mode, and provided the same read alignments from bowtie2 or SHRiMP which were used for the other read tallying approaches. In addition, we used the dedicated tRNA-Seq quantification tool, mimseq, which performs a bespoke alignment and quantification.

With the exception of Decision and mimseq, which inherently generate quantification at multiple levels, all other approaches quantify at the level of the Gene locus ID sequences included in the reference fasta. We further summarised quantification to transcript ID and anticodon-level feature abundances by summation. We also summarised the quantification values to the same ‘mimseq isodecoder’ features quantified by mimseq, to make them comparable. A summary of the 4 levels of quantification and how these are obtained for each read tallying approach is shown (Table 2).

The proportion of reads used by each read tallying approach varies considerably. Salmon, Fractional and Random use all mapped reads for each level of quantification, whereas stricter approaches which filter by multimapping or mapping quality use the least reads and frequently assign fewer than 50% of the mapped reads (Figure 4a; Supplementary Figure 3a). As intended, Decision uses increasingly more reads at higher levels of quantification. Importantly, Decision uses approximately the same number of reads as mimseq at the mimseq isodecoder level, but consistently more at the anticodon level, where it approaches the use of all mapped reads. In contrast, Mimseq uses the same number of reads for both mimseq isodecoder and anticodon level quantification. This suggests Decision is more appropriately determining which reads to utilise for each level of quantification.

a) Reads assigned at anticodon and mimseq isodecoder level for each read tally method. Only results from alignment with bowtie2 or GSNAP (Mimseq) are shown. b) Mean RMSE at anticodon and mimseq isodecoder level. *=Best approach for each tRNA-Seq method, ▾=Worst approach. c). As per (b), except Pearson correlation coefficient. d) The percentage of reads from each simulation sample which are single or multiple aligned, separated by whether the alignment position(s) have the correct anticodon. Multi-aligned reads with more than more anticodon were deemed incorrect.

Comparing read tallying approaches

To assess the overall accuracy of the abundance estimates obtained with the varying read tallying approaches, we calculated two metrics. The root mean squared error (RMSE) is a common error metric which captures how closely the quantification estimates match the ground truth31. However, in many applications of tRNA-Seq, including the comparison between conditions or tissues, it’s sufficient to accurately capture the abundance relationship between samples, even if there is a systematic over or under-estimation across all samples. Hence, we complemented the evaluation using RMSE by also considering the Pearson correlation coefficient between the quantification estimates and ground truth.

To compare the quantification approaches, we considered the mean RMSE and mean correlation across all features at a given quantification level. At the anticodon and mimseq isodecoder level, our novel approaches, Decision and Salmon consistently show the lowest RMSE and highest correlation, suggesting they deal with multi-mapped reads most appropriately. Of the previously used read-tallying approaches, Mimseq shows the highest accuracy. Surprisingly, Random and Fractional also perform well at the anticodon level, while Unique and MAPQ>10 performed poorly (Figure 4b-c). On further inspection, this is because the multi-mapped reads are predominantly mapped to multiple tRNA species from the same anticodon (Figure 4d; Supplementary Figure 3a). Thus, requiring unique alignments at the level of individual tRNA sequences (using Unique or MAPQ>10) leads to detrimental discarding of multi-mapped reads which could be assigned to a single anticodon. For these reads, selecting one alignment at random, or equally distributing the read between all of them, is a more sensible approach and thus leads to more accurate anticodon level quantification. However, for quantification at lower levels, Unique and MAPQ > 10 are more accurate than Random and Fractional, suggesting discarding of multi-mapped reads is better than naive resolution for quantification below the anticodon level (Figure 4b-c). Very similar results are observed when SHRiMP is used in place of bowtie2 (Supplementary Figure 3b-c). Finally, for quantification at the level of individual tRNA sequences, all methods perform equally well, with the exception of Random, which performed the worst. Overall, Decision and Salmon are consistently the best read tallying methods to apply.

We reasoned that the misalignment of reads should be one of the major factors reducing quantification accuracy for all methods. In support of this, the fraction of reads correctly aligned and the quantification evaluation metrics were correlated (Supplementary Figure 4a-b). The exception was MAPQ>10 at anticodon level, which shows poor accuracy even when the proportion of correctly aligned reads is high. This agrees with the previous observation that discarding low quality reads is detrimental for anticodon-level read tallying because reads are typically mapped across multiple tRNAs for the same anticodon, not between anticodons (Figure 3d). Thus, excepting quantification by MAPQ>10, our results indicate that misalignment is one of the major issues with tRNA-Seq quantification.

Using simulations to inform interpretation of results

Summarising the RMSE and correlation metrics to the mean value over all features enables read tallying approaches to be easily compared, but masks fine-grained differences which go against the trend and could give false confidence when interpreting quantification of individual features. We therefore further considered which features each approach quantifies best, focusing on the simulations from the YAMAT-Seq data and quantification at the anticodon level. For these simulations, Decision had the highest mean correlation and lowest mean RMSE, with Salmon performing nearly equivalently and Mimseq also being accurate (Figure 4c). However, we observed that the correlations for some anticodons were very sensitive to the read tallying approach (Figure 5a-b). For example, the correlation for Ser-GGA is much lower with Mimseq than Decision or Salmon, whilst the correlation for Pro-AGG is much higher with Mimseq than Decision or Salmon (Figure 5b; Supplementary Figure 4c). This suggests the choice of quantification approach may have different impacts on quantification accuracy for each anticodon and the interpretation of results should ideally take into account the limits of each quantification approach.

Each quantification approach has specific shortfalls

a) Pearson correlation between ground truth and estimated read counts for six anticodons with variable quantification accuracy between read tallying approaches. b). c) The variance in observed fold-changes in the real tRNA-Seq data that is explained by fold-changes in the uniform simulation data. x = The mean variance explained across the four Homo sapiens datasets.

Given the observed association between read misassignment and quantification accuracy, we hypothesised that differences in read misalignments between cell lines could erroneously induce apparent differences in tRNA feature abundances. To test this, we compared the fold-change between cell lines when simulating the same number of reads for each tRNA with the fold-changes observed in the real data. The fold-changes with the simulated data should reflect technical biases resulting from systematic differences in misassignments between cell lines due to differences in the error profiles. Strikingly, up to 13% of the variance in fold-changes in the real data were explained by technical biases (Figure 5c). Even using the most accurate quantification method, Decision, up to 7% of the variance for anticodon level fold-changes in the real data is explained by technical biases. All quantification methods appeared to show some propensity to identify fold-changes between conditions that were partially explained by technical biases for at least one level of tRNA quantification. The exception to this was salmon, for which a maximum of 2% of the variance in fold-changes were explained by technical biases. This suggests salmon may be a more trustable quantification method for the detection of fold-changes between experimental conditions.

Discussion

Recent advances in sequencing using nanopores promise a potential revolution for RNA sequencing, including tRNAs32. Despite this, illumina sequencing is expected to remain the dominant approach for tRNA sequencing, at least in the short-term. The first tRNA-seq approaches using illumina sequencing were published in 201514,15, since which, the comparison of bioinformatic approaches for tRNA quantification from tRNA-Seq has received surprisingly little attention. In part, this may be because tRNA-Seq methods continue to be developed and the optimal quantification is considered to be dependent upon the method. Encouragingly, however, we have identified two robust quantification methods that appear to be suitable for all tRNA-Seq data.

An important aspect of our benchmarking is the simulation of realistic tRNA-Seq data, which has not been attempted before, to our knowledge. The only previous attempt to simulate tRNA-Seq reads that we are aware of did not capture the effect of modifications on read truncations and mutations33. By learning error profiles of individual samples and then simulating data from these error profiles, our simulations at least partially capture the technical variability between samples, cell lines and tRNA-Seq methods. However, there are limitations to this approach. All truncation and misincorporation events are measured independently, such that the error profiles will not capture dependencies between the events, which occur with respect to correlated misincorporations at nucleotides due to linked modifications34. In addition, the error profiles are inherently biased by the ability to align the tRNA-Seq reads. Misincorporations and truncations which prevent the read being aligned appropriately will not be captured in the error profiles. As such, the simulated data may be considered to simulate the alignable portion of a tRNA-Seq sample, with biases due to the alignment strategy used. Here, we attempted to mitigate this by aligning reads with multiple aligners. Despite these limitations, our simulations represent the most realistic simulation of tRNA-Seq data to date, and enabled us to perform a rigorous comparison of tRNA-Seq quantification approaches.

The crucial steps for accurate tRNA-Seq quantification are the alignment of reads and handling of multi-mapped reads, with the latter being the step which varies most between the approaches that have been applied to date. Although reads may be multi-mapped across different tRNA species, we show here that the issue is most pressing at the Gene locus ID and Transcript ID level, and much less so at the Anticodon level. However, dealing with multi-mapped reads inappropriately can itself create issues at the anticodon level. The approaches used to date resolve the multi-mapping are only suitable for quantification at particular levels, which has not previously been appreciated. Crucially, the common approach of removing multi-mapped reads and/or low quality reads is significantly suboptimal for the anticodon level quantification which it has been previously employed for12,14, since it leads to the unnecessary discarding of reads which have multi-mapped across tRNAs with the same anticodon. The bespoke quantification approach of Mimseq is better than filtering multi-mapped reads, but our simulations indicate it performed no better than simple random assignment of multi-mapped reads for anticodon-level quantification.

The central observation of this benchmarking study is that two simple approaches alleviate the issue of multi-mapped reads and perform better than existing approaches at all levels of quantification. This is observed across all 4 of the tRNA-Seq methods, suggesting our observation can likely be extrapolated to all tRNA-Seq methods using illumina sequencing for full length tRNAs. The Salmon-based approach involves reporting all alignment positions and then using salmon to probabilistically resolve the multi-mapped reads. The Decision approach involves counting reads at each level of quantification where the read maps to tRNAs from a single feature, e.g multiple tRNAs from the same anticodon. Given the lower propensity for technical biases to explain fold-changes in real data with Salmon and the reduced barrier to its use, we suspect salmon will prove to be the most popular. For this approach, bowtie2 simply needs to be run with the -a flag to report all reads. The alignments must then be filtered to retain only the equal top-scoring alignments for each read, before passing the alignments to salmon quant.

While the motivation of this study was to determine the optimal overall quantification approach, our simulations also identified specific tRNAs that cannot be accurately quantified in particular samples. Our simulations further allowed us to establish that a small proportion of the observed variability between cell lines may not be biological. Thus, comparisons between samples should be interpreted with caution, since they may be partially due to technical biases in read miss-alignment. We propose that a simulation-based approach could be applied more widely as a quality control step to highlight features whose quantification may be inaccurate. In particular, where changes in tRNA abundance between cell lines are observed, it would be prudent to use similar simulations to demonstrate that such changes are not driven by differences in tRNA modifications driving systematic differences in misincorporations and truncations.

The decision about what level to interrogate tRNA abundance is a trade off between a desire to obtain the most ‘fine-grained’ data possible, and the practical consideration of how accurately tRNAs can be quantified at each level. tRNA isodecoders have been found to be functionality distinct35,36 and differential abundance of specific Transcript IDs has been identified across mouse tissues12 and human cell lines16. Nonetheless, tRNA abundances are more accurately quantified at the Anticodon level and this is likely to remain the predominant approach for the foreseeable future. As tRNA-Seq methods continue to improve the processivity of reverse transcriptase over modified bases and yield fewer truncated reads, the potential for highly accurate Transcript ID level quantification will open up. Sequencing tRNAs using nanopores is also likely to provide a significant benefit in this respect32. However, these improvements in experimental methods need to be complemented by rigorous improvements in the data processing to take full advantage of the potential benefits. We envisage the simulation-based approach we have used here being reutilised as new experimental methods are developed, to maintain a high standard of tRNA-Seq data analysis.